---
title: "Harmonization of a multi-site imaging dataset using neuroCombat"
author: 
- name: Jean-Philippe Fortin
  email: fortin946@gmail.com
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc_float: true
#    theme: paper
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Harmonization of a multi-site imaging dataset using neuroCombat}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
bibliography: references.bib
---


# Intro

Vignette describing a function in development, ```neuroCombatFromTraining```.  The function applies ComBat scanner correction factors estimated from a training dataset to a test dataset. Scanners/sites of the test dataset must be included in the training dataset. 

# Example dataset


```{r}
library(neuroCombat)
library(neuroCombatData)
se <- seCorticalThickness
se
```

We use the example dataset stored in the ```neuroCombatData``` package: 10 scans from 2 scanners, with 200 imaging features. Scans come from either a normal or disease brain.  


# Harmonization

```{r}
data <- assays(se)[["freesurfer"]]
mod <- model.matrix(~colData(se)$gender+colData(se)$age)
results <- neuroCombat(dat=data,
                       batch=colData(se)$site,
                       mod=mod)
```

```{r}
names(results)
```

```{r}
neuroCombat::drawPriorDelta(results$estimates, xlim=c(0,2.5))
neuroCombat::drawPriorGamma(results$estimates, xlim=c(-1.5,1.5))
```



  

```{r}
data_standardized <- results[["dat.standardized"]]
col <- as.numeric(as.factor(colData(se)$site))
boxplot(data_standardized, ylim=c(-2,2),col=col)
```


```{r}
data_norm <- results[["dat.combat"]]
data_norm <- standardizeData(data_norm,
                             batch=colData(se)$site,
                             mod=mod)
boxplot(data_norm, ylim=c(-2,2),col=col)
```




# References

