---
title: "Harmonization of a test dataset using ComBat parameters estimated from a training dataset "
author: 
- name: Jean-Philippe Fortin
  email: fortin946@gmail.com
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc_float: true
#    theme: paper
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Removing scanner effects in imaging data with neuroCombat}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
bibliography: references.bib
---


# Intro

Vignette describing a function in development, ```neuroCombatFromTraining```.  The function applies ComBat scanner correction factors estimated from a training dataset to a test dataset. Scanners/sites of the test dataset must be included in the training dataset. 

# Example dataset

We use the example dataset stored in the ```neuroCombat``` package: 10 scans from 2 scanners, with 200 imaging features. Scans come from either a normal or disease brain.  

```{r, warning=FALSE, message=FALSE}
library(neuroCombat)
data  <- combatExampleData
batch <- combatExampleBatch
pheno <- combatExamplePheno
mod   <- model.matrix(~pheno)
batch
pheno
head(data)
```

# Testing if the code works as intended 

Training: full dataset (10 scans)
Test: subset of the full dataset (4 scans, 2 from each scanner)

```{r}
indices.training <- c(1:10)
indices.test     <- c(1,2,6,7)
data.training    <- data[, indices.training]
data.test        <- data[, indices.test]
batch.training   <- batch[indices.training]
batch.test       <- batch[indices.test]
```

## Scenario 1: no covariates adjustment in training dataset and no covariates adjustment in test dataset

We estimate scanner effects using the full dataset as training dataset, and apply the scanner correction on the test dataset. We don't apply any covariates adjustment. Since the test dataset is part of the training dataset, we should obtain identical harmonized data. 

```{r, warning=FALSE, message=FALSE}
data.training.harmonized <- neuroCombat(data.training, batch=batch.training)
data.test.harmonized <- neuroCombatFromTraining(data.test, batch=batch.test, estimates=data.training.harmonized$estimates)
```

```{r, echo=FALSE}
col <- c(1,1,2,2)
ylim=c(-5,20)
par(mfrow=c(1,3))
boxplot(data.training[, indices.test],ylim=ylim, las=2,col=col, main="Un-harmonized")
boxplot(data.training.harmonized$dat.combat[, indices.test],ylim=ylim, las=2, col=col, main="Harmonized (training + test)")
boxplot(data.test.harmonized, ylim=ylim, las=2, main="Harmonized (test only)", col=col)
```

```{r, echo=FALSE, fig.height=3}
scans <- colnames(data)[indices.test]
par(mfrow=c(1,4))
X <- data.training.harmonized$dat.combat[, indices.test]
Y <- data.test.harmonized
for (i in 1:4){
  plot(X[,i], Y[,i], main=scans[i],
       xlab="Values from training + test",
       ylab="Values from test only")
  abline(a=0,b=1)
}
```

Alright, it seems to be working as intended. 

## Scenario 2: covariates adjustment in training dataset, but no covariates adjustment in test dataset

In this scenario, since no covariates are provided in the test dataset, we expect the harmonized data from the training and test datasets to differ. This happens because the internal data standardization prior to scanner correction in the ComBat algorithm takes into account covariates. 

To standardize the test dataset, we use an average of the standardization factors from the training dataset.  


```{r, warning=FALSE, message=FALSE}
data.training.harmonized <- neuroCombat(data.training, batch=batch.training, mod=mod)
data.test.harmonized <- neuroCombatFromTraining(data.test, batch=batch.test, estimates=data.training.harmonized$estimates)
```

```{r, echo=FALSE}
col <- c(1,1,2,2)
ylim=c(-5,20)
par(mfrow=c(1,3))
boxplot(data.training[, indices.test],ylim=ylim, las=2,col=col, main="Un-harmonized")
boxplot(data.training.harmonized$dat.combat[, indices.test],ylim=ylim, las=2, col=col, main="Harmonized (training + test)")
boxplot(data.test.harmonized, ylim=ylim, las=2, main="Harmonized (test only)", col=col)
```

```{r, echo=FALSE, fig.height=3}
scans <- colnames(data)[indices.test]
par(mfrow=c(1,4))
X <- data.training.harmonized$dat.combat[, indices.test]
Y <- data.test.harmonized
for (i in 1:4){
  plot(X[,i], Y[,i], main=scans[i],
       xlab="Values from training + test",
       ylab="Values from test only")
  abline(a=0,b=1)
}
```

As expected, there are subtle differences between the training and test datasets, but overall the scanner correction appears to be successful. 

## Scenario 3: covariates adjustment in training dataset, and covariates adjustment in test dataset

Not supported yet. 

# References



